The role of relevance, competence and priors for scalar implicatures

### 2022-04-17 writing/ELM2/abstract.tex
fix analyses and plots with correctly z-scored data

\documentclass[11pt,letterpaper]{article}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{helvet}
\pagestyle{empty}
\usepackage[kerning=true]{microtype}
\usepackage{parskip}
\usepackage{sansmath}
\usepackage{graphicx}
\usepackage{sidecap}  
\sidecaptionvpos{figure}{c}
\usepackage{float}
\usepackage{color, soul}
% Feel free to use additional packages for glosses, figures, whatnot.
\usepackage[dvipsnames]{xcolor}
\newcommand{\mf}[1]{\textcolor{BurntOrange}{[MF: #1]}}
\newcommand{\pt}[1]{\textcolor{Cerulean}{[PT: #1]}}
\newcommand{\bvt}[1]{\textcolor{ForestGreen}{[BvT: #1]}}
% The next bit is for reserving sufficient space for authors,
% affiliations, and e-mail address.  No need to change for initial
% anonymous version.  For the final version, replace the
%\toggletrue{anonymous} %with \togglefalse{anonymous} to de-anonymize.
\usepackage{etoolbox}
\newtoggle{anonymous}
\toggletrue{anonymous}

@@

\title{The role of relevance, competence and priors for scalar implicatures}
\authors{Michael Franke (University of T\"ubingen), Bob van Tiel (Radboud University), Polina Tsvilodub (Osnabr\"uck university)}
\email{pstvilodub@uos.de}

% Here goes the main text of your abstract:

If someone says ``Anna ate some cookies'', the hearer might infer the upper-bounded reading that Anna ate \textit{some, but not all} cookies. Similarly, given ``Donald ate a donut or a pretzel.'', one might infer that Donald ate either \textit{the donut or the pretzel, but not both} (i.e., an exclusive interpretation). 
Both inferences are usually explained as a variety of \textit{scalar implicature} (SI). SIs rely on lexical scales consisting of words ordered in terms of informativeness, like $\langle$some, all$\rangle$ and $\langle$or, and$\rangle$. If a speaker uses an informationally weaker term (e.g., ``some''), they may imply that the corresponding stronger alternative (e.g., ``all'') is false [3,5]. %: if the stronger alternative was true, a rational agent would have used the stronger utterance; but since she did not, she must have lacked the evidence for using ``all'' 
%(Grice, 1975).
%Some prior work addresses the influence of sentence structure, or of specific linguistic markers on the robustness of SIs (Li, 2021). Moreover, 
Crucially, prior research suggests that the robustness of SIs is influenced by different contextual factors [1].
We investigate the effects of three factors: (1) the \textit{competence} of the speaker about the truth of the stronger alternative, (2) its \textit{relevance} to the listener, and (3) the \textit{prior probability} that it is true [2,4,6]. We explore how these three factors \textit{interactively} influence the robustness of SIs associated with the triggers ``some'' and ``or''.

@@

%(71  excluded following preregistered exclusion criteria).
We analysed data from 206 participants recruited on Prolific. Their ratings were z-scored within each factor by-participant. 
We regressed the implicature likelihood ratings against the fixed effects of trigger, all factor ratings within-story, all interactions and maximal  random effects, using a Bayesian linear mixed effects model. 
%We analysed the results using a Bayesian linear mixed effects model, regressing the implicature likelihood rating against the fixed effects of all factor ratings within-story, the effect of trigger, all interactions and maximal tractable random effects. 
Participants' factor ratings by-story agreed well with the designed classification of the stories (Fig.~\ref{by-item-ratings}, red~vs.~blue color on x-axis).
%Consistent with predictions of the SI account, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true (probability of the effect of prior being smaller than 0.05: $P = 0.999$). 
As predicted, participants were more likely to derive the SIs of ``some'' and ``or'' when judging speaker competence as high ($P=0.999$ for ``some'',  $P=0.993$ for ``or'' for effect sizes being $>0.05$, Fig.~\ref{posteriors} for all results). 
They were also more likely to derive the SI of ``some'' when judging the prior probability of ``all'' as low ($P= 1$ for effect $<0.05$), which was not the case for ``or''. %Interestingly, we did not observe a credible effect of prior probability in the case of ``or''.
Finally, we did not observe credible effects of relevance for either trigger.
%For the trigger ``or'', we only find a positive effect of competence ($P =  0.993$). 
Given the unexpected absence of prior effects for ``or'', we computed exploratory pairwise correlations of all predictors. While no correlations were found for ``some'', we found a significant correlation between the explanatory factors prior and relevance for ``or'' ($R^2=-0.106, p<0.01$). %, and a significant correlation between competence and relevance ($R^2 = 0.127, p=0.001$) for ``or''. 
An exploratory model comparison of two models, one containing the relevance effect over one containing the prior effect, each combined with competence, revealed mild evidence in favor of the prior as a better explanatory factor than relevance (Bayes Factor $ = 3.70$). %This might be due to the difficulty of accessing relevance experimentally.
While supporting the view that SIs for both triggers rely on epistemic reasoning affected by speaker competence, our results indicate that prior and relevance might be closely connected for ``or''.
Ultimately, our results suggest that the interdependence of the three factors is more complex than just the sum of the effects anticipated in the literature, and provide further insights into how exactly they might enter into people’s decision as to whether or not to derive an SI.

>>>>

\documentclass[11pt,letterpaper]{article}
\usepackage[margin=1in]{geometry}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{hyperref}
\renewcommand{\familydefault}{\sfdefault}
\usepackage{helvet}
\pagestyle{empty}
\usepackage[kerning=true]{microtype}
\usepackage{parskip}
\usepackage{sansmath}
\usepackage{graphicx}
\usepackage{sidecap}  
\sidecaptionvpos{figure}{c}
\usepackage{float}
\usepackage{color, soul}
% Feel free to use additional packages for glosses, figures, whatnot.
\usepackage[dvipsnames]{xcolor}
\newcommand{\mf}[1]{\textcolor{BurntOrange}{[MF: #1]}}
\newcommand{\pt}[1]{\textcolor{Cerulean}{[PT: #1]}}
\newcommand{\bvt}[1]{\textcolor{ForestGreen}{[BvT: #1]}}
% The next bit is for reserving sufficient space for authors,
% affiliations, and e-mail address.  No need to change for initial
% anonymous version.  For the final version, replace the
%\toggletrue{anonymous} %with \togglefalse{anonymous} to de-anonymize.
\usepackage{etoolbox}
\newtoggle{anonymous}
\togglefalse{anonymous}

@@

\title{The role of relevance, competence and priors for scalar implicatures}
\authors{Polina Tsvilodub (Osnabr\"uck University, ptsvilodub@uos.de), Bob van Tiel (Radboud University), Michael Franke (University of T\"ubingen)}
%\email{}
% Here goes the main text of your abstract:
If someone says ``Anna ate some cookies'', the hearer might infer the upper-bounded reading that Anna ate \textit{some, but not all} cookies. Similarly, given ``Donald ate a donut or a pretzel.'', one might infer that Donald ate either \textit{the donut or the pretzel, but not both} (i.e., an exclusive interpretation). 
Both inferences are usually explained as a variety of \textit{scalar implicature} (SI). SIs rely on lexical scales consisting of words ordered in terms of informativeness, like $\langle$some, all$\rangle$ and $\langle$or, and$\rangle$. If a speaker uses an informationally weaker term (e.g., ``some''), they may imply that the corresponding stronger alternative (e.g., ``all'') is false [3,5]. %: if the stronger alternative was true, a rational agent would have used the stronger utterance; but since she did not, she must have lacked the evidence for using ``all'' 
%(Grice, 1975).
%Some prior work addresses the influence of sentence structure, or of specific linguistic markers on the robustness of SIs (Li, 2021). Moreover, 
Crucially, prior research suggests that the robustness of SIs is influenced by different contextual factors [1].
We investigate the effects of three factors: (1) the \textit{competence} of the speaker about the truth of the stronger alternative, (2) its \textit{relevance} to the listener, and (3) the \textit{prior probability} that it is true [2,4,6]. We explore how these three factors \textit{interactively} influence the robustness of SIs associated with the triggers ``some'' and ``or''.

@@

%(71  excluded following preregistered exclusion criteria).
We analysed data from 206 participants recruited on Prolific. Their ratings were z-scored within each factor by-participant. 
We regressed the implicature likelihood ratings against the fixed effects of trigger, all factor ratings within-story, all interactions and maximal  random effects, using a Bayesian linear mixed effects model. 
%We analysed the results using a Bayesian linear mixed effects model, regressing the implicature likelihood rating against the fixed effects of all factor ratings within-story, the effect of trigger, all interactions and maximal tractable random effects. 
Participants' factor ratings by-story agreed well with the designed classification of the stories (Fig.~\ref{by-item-ratings}, red~vs.~blue color on x-axis).
%Consistent with predictions of the SI account, for the trigger ``some'', we find a clear negative prior probabilty effect of the stronger alternative ``all'' being true (probability of the effect of prior being smaller than 0.05: $P = 0.999$). 
As predicted, participants were more likely to derive the SIs of ``some'' and ``or'' when judging speaker competence as high ($P=0.999$ for ``some'',  $P=0.993$ for ``or'' for effect sizes being $>0.05$, Fig.~\ref{posteriors} for all results). 
They were also more likely to derive the SI of ``some'' when judging the prior probability of ``all'' as low ($P= 1$ for effect $<0.05$), which was not the case for ``or''. %Interestingly, we did not observe a credible effect of prior probability in the case of ``or''.
Finally, we did not observe credible effects of relevance for either trigger.
%For the trigger ``or'', we only find a positive effect of competence ($P =  0.993$). 
Given the unexpected absence of prior effects for ``or'', we computed exploratory pairwise correlations of all predictors. While no correlations were found for ``some'', we found a significant correlation between the explanatory factors prior and relevance for ``or'' ($R^2=-0.106, p<0.01$). %, and a significant correlation between competence and relevance ($R^2 = 0.127, p=0.001$) for ``or''. 
An exploratory model comparison of two models, one containing the relevance effect over one containing the prior effect, each combined with competence, revealed mild evidence in favor of the prior as a better explanatory factor than relevance (Bayes Factor $ = 3.70$). %This might be due to the difficulty of accessing relevance experimentally.
While supporting the view that SIs for both triggers rely on epistemic reasoning affected by speaker competence, our results indicate that prior and relevance might be closely connected for ``or''.
Ultimately, our results suggest that the interdependence of the three factors is more complex than just the sum of the effects anticipated in the literature, and provide further insights into how they might enter into people’s decision as to whether or not to derive an SI.
